Aa Language Reference Manual

### 2010-07-06 Ahir-asm/antlr2/docs/AaLRM.tex
Some more changes to the grammar to rationalize the syntax.  Introduced a $place statement to enable branches.  Also started adding link code in the parsing phase.

The {\bf Aa} language is intended to be used to
describe an algorithm so that the control flow
in the description can be modeled by a petri net
of a specific class (the type-2 petri-net as
introduced in \cite{ref:SameerPhD}).  A program in
\Aa is also a description of a system which reacts with 
its environment through input and output ports and 
through message queues.

@@

The module body itself is a sequence of statements.
\begin{verbatim}
module-body := (statement)+
\end{verbatim}
The control flow implied is serial in nature.  The
module starts executing when it gets a token.
This token is passed to the first statement in the
list which executes and releases the token when it
finishes execution, and the token  is passed serially down the
list of statements.  When the last statement finishes 
executing, the token passes out of the module.

@@

The control flow in an \Aa module can be visualized
in terms of the flow of tokens.
A module is invoked when it receives a token, and 
this token flows down the sequence of statements.
Whenever a statement receives a token, it executes,
and when it has finished executing, the token is
released to the next statement.

@@

Statements in the module-body can be simple statements
or block statements.

@@

@@

restriction is that this directed graph must be acyclic!

@@

           ( expr-value  statement-to-be-executed )*
           ( default statement-to-be-executed )?

\end{verbatim}
The switch-expression is checked, and depending on its
value, one of the alternative statements is selected.
Thus, the incoming token to the switch statement is 
passed to one of the alternatives.  For example,
\begin{verbatim}
$switch a $when 0 $then a1 := (b + c)
          $when 1 $then a2 := (b - c)
          $default $null
$endswitch
\end{verbatim}
The control token is routed to the appropriate
choice and if the token is not routed out 
of the block (by the {\bf takeroute} statement 
described below), then the token passes to the
statement following the switch. 
\item If statements: An if statement has the form
\begin{verbatim}
if-statement :=
       if-spec test-expression 
           true-statement
       (else
           false-statement)?

\end{verbatim}
For example, 
\begin{verbatim}
$if (a != 0) $then
   q := (r + s)
   t := 0
$else
   qdash :  (r - s)
$endif
\end{verbatim}
\item Routing statement: The {\bf takeroute} statement
defines a labeled path along which it sends the
incoming control token.  For example,
\begin{verbatim}
$takeroute [fastpath]
\end{verbatim}
means that the incoming token is not sent out to the
next statement in the sequence, but is sent out along
a path labeled ``fastpath''.  The token will be captured
either by a {\bf merge} statement, or will exit the
containing branch block.
\item Merge statements: A merge statement is
specified as 
\begin{verbatim}
merge-statement :=
          merge-spec label-list merge-assignments
\end{verbatim}
which is to be interpreted as follows.  The labels in the label list
refer to paths defined by ``takeroute'' statements within the branch block.
Whenever a token arrives along any of these paths,
the merge statement starts and executes a series of special assignments
which multiplex values into new variables based on which arc the
token arrived from.  The merge statement then releases its token to
the next statement. 
\end{itemize}

Here is an example of a branch block constructed in this manner
\begin{verbatim}
$branchblock [b1] 
{
  $merge loop $entry // $entry is the path 
                     // into the block

       $phi q := 0 on $entry r on loop 
                    // q is defined by where 
                    // the token came from
  r := q + 1
  $if r < 10 $then 
       $takeroute [loop]
  $endif
}
\end{verbatim}

\end{itemize}

Thus, the program is constructed as a collection of modules,
each of which is a sequence of statements.  The use of series,
parallel, fork and branch block
statements enables the programmer to describe a highly concurrent
structured system with complex branching behaviour.  
The resulting control flow structure is a
type-2 petri-net with provable liveness and safety properties \cite{ref:SameerPhD}.

\section{Scoping Rules}

@@

\section{Syntax}

The syntax for \Aa is described by the following lexical analysis and
parsing rules (written as rules to be parsed by antlr2 \cite{ref:antlr2},
a LL(k) parser).

This is the set of tokens recognized by the lexical analyzer (or lexer).
\input{AaLexer}
As one can see, all keywords in the \Aa language begin with \$.

The grammar for the parser is (using the EBNF notation) given below. 
\input{AaParser}

@@

A little bit more complicated:
\begin{verbatim}
// an array of 32-bit unsigned integers.
$storage mem $array<1024> of $uint<32>

// module returns the sum of mem[I] from
// I=low to I=high
$module [sum_mod]
    $in (low $uint<10> high $uint<10>)
    $out (sum $uint<32>)
$is
    $storage I $uint<10>
    $storage J $uint<10>
    $storage sum1 $uint<32>
    $storage sum2 $uint<32>
{

@@

   I := low
   J := (mp + 1)

   \branchblock[trivcheck]
    {
       $merge $entry
       $if (d > 0) $then
           // do two summations in parallel
           // parallel summations
           \parallelblock[parsum]
           {
               \branchblock[sb1]{
                    merge $entry loopback
                       $phi s := 0 on $entry %forward:s on loopback
                    $if (I < mp) $then
                        \seriesblock[loopback]{
                            I := (I+1)
                            s := $mux (I == 0)  0 (s + mem[I])
                         }
               } 
               \branchblock[sb2]{
                    merge $entry loopback
                       $phi s := 0 on $entry %forward:s on loopback
                    $if (J < high) $then
                        \seriesblock[loopback]{
                            J := (J+1)
                            s := $mux (J == (mp+1))  0 (s + mem[J])
                         }
               } 
           }

           // combine results from parallel statement above
           snontriv := (%parsum%sb1:s  + %parsum%sb2:s)
           $takeroute [nontrivsum]
      $else

           // summation is trivial 
           striv := mem[low]
           $takeroute [trivsum]
      $merge nontrivsum trivsum
         // which sum do you pick? depends on which path was taken
         $phi sum := snontriv on nontrivsum strivsum on trivsum
      $exit
    }

    // bring sum up to the surface.
    sum := %trivcheck:sum
}
\end{verbatim}

>>>>

{\bf Aa} is intended to be a low-level 
programming language for the description
of algorithms in which the control flow
can be modeled by a petri net
of a specific class (the type-2 petri-net as
introduced in \cite{ref:SameerPhD}).  A program in
\Aa can also be viewed as a 
description of a system which reacts with 
its environment through input and output ports and 
through message queues.  Thus, an \Aa program
can be directly executed on a computer, and
can also be mapped to a logic circuit.

@@

The module body itself is a sequence of statements.
\begin{verbatim}
module-body := (statement)+
\end{verbatim}
Each statement can be viewed as a transition
in a petri-net, and has a 
set of input (or source) places  and a set of output
(or sink)  places.  
The execution of the statement is triggered when
there is a token present in each of its source 
places and when the statement finishes execution,
a token is placed in each of its output places.
The control flow in the sequence of statements
in a module is serial in nature.  Thus each
statement has a single source place and a single
sink place, sink place of a statement being 
the source place of its successor.
The module as a whole can be viewed as having
a single entry place and a single exit place.
When a token appears in the entry place, the first
statement can trigger, and when the last statement
has finished, a token is placed in the exit
place.

@@

@@

Statements in the module-body are can be of two kinds:
\begin{itemize}
\item Atomic statements: those that have a single
entry place and a single exit place.  These are
the only statements that can occur in a module body.
These statements are further divided into 
{\em simple} statements and {\em block} statements.
\item Sub-atomic statements:  these statements
can appear only inside block statements,  and they
can have multiple sources and multiple sinks.
Examples of these statements are statements for
combining tokens (merge and join statements) and
statements for redirecting tokens (the toss statement).
\end{itemize}

@@

                sequence-of-atomic-statements

@@

restriction is that this directed graph must be acyclic!  This is enforced
by requiring that a join statement refers only to labels of statements
that appear before the join.

@@

           ( expr-value  serial-statement-sequence )*
           ( default serial-statement-sequence )?

\end{verbatim}
The switch-expression is checked, and depending on its
value, one of the alternative statements is selected.
Thus, the incoming token to the switch statement is 
passed to one of the alternatives.  For example,
\begin{verbatim}
$switch a $when 0 $then a1 := (b + c)
          $when 1 $then a2 := (b - c)
          $default $null
$endswitch
\end{verbatim}
The control token is routed to the appropriate
choice sequence and if the token is not routed out 
of the block (by the {\bf place} statement 
described below), then the token passes to the
statement following the switch. 
\item If statements: An if statement has the form
\begin{verbatim}
if-statement :=
       if-spec test-expression 
           serial-statement-sequence
       (else
           serial-statement-sequence)?

\end{verbatim}
For example:
\begin{verbatim}
$if (a != 0) $then
   q := (r + s)
   t := 0
$else
   qdash :  (r - s)
$endif
\end{verbatim}
If the control token reaches the end of a selected
segment in the if statement (that is, without being
rerouted by a place statement), then the control
token is passed on to the statement immediately
following the if statement.
\item The place statement: The {\bf place} statement
defines a place with a label  into which it places
a token after it has executed.  For example,
\begin{verbatim}
$place [fastpath]
\end{verbatim}
means that the incoming token is placed in a labeled
place ``fastpath''.  The token will either
trigger a {\bf merge} statement which is required
to depend on this labeled place in the same branch
block, or if no such merge statement
exists, then ``fastpath'' is identified
with the exit place of the  containing branch block.
\item Merge statements: A merge statement is
specified as 
\begin{verbatim}
merge-statement :=
          merge-spec label-list merge-assignments
\end{verbatim}
which is to be interpreted as follows.  The labels in the label list
refer to token labels defined by ``toss'' statements within the branch block.
Whenever a token labeled by any of the items in label-list arrives
at the input of the merge,
the merge statement starts and executes a series of special assignments
which multiplex values into new variables based on which arc the
token arrived from.  The merge statement then releases its token to
the next statement. 
\end{itemize}

Here is an example of a branch block constructed in this manner
\begin{verbatim}
$branchblock [b1] 
{
  $merge loop $entry // $entry is the place
                     // into the block
                     // merge triggered by
                     // token at entry or
                     // by token in place "loop"

       $phi q := 0 on $entry r on loop 
                    // q is defined by where 
                    // the token came from
  r := q + 1
  $if r < 10 $then 
       $place [loop] // put token in place "loop"
  $endif
}
\end{verbatim}
This is to be interpreted as follows: the merge executes
whenever a token is present  either in the entry place
of the branch-block or in the place labeled ``loop''.
The merge defines variable ``q'' with a value which is
either 0 (if a token was present in the entry place)
or r (if a token was present in the ``loop'' place).
By the construction rules in an \Aa program, it is
impossible for there to be a token present in more
than one input place to a merge.
\end{itemize}

Thus, an \Aa program is constructed as a collection of modules,
each of which is a sequence of statements.  The use of series,
parallel, fork and branch block
statements enables the programmer to describe a highly concurrent
structured system with complex branching behaviour.  
The resulting control flow structure is a
type-2 petri-net with provable liveness and safety properties \cite{ref:SameerPhD}.

\section{Scoping Rules} \label{sec:Scopes}

@@

@@

An example which is a little bit more complicated:
\begin{verbatim}
// an array of 32-bit unsigned integers.
$declare $storage mem $array<1024> $of $uint<32>

// module returns the sum of mem[I] from
// I=low to I=high
$module [sum_mod]
    $in (low $uint<10> high $uint<10>)
    $out (sum $uint<32>)
$is
{

@@

   $branchblock[trivcheck]
    {
       // d from parent scope
       $if (d > 0) $then  
           // do two summations in parallel
           // parallel summations
           $parallelblock[parsum]
           {
               $branchblock[sb1]{
                    $declare $storage  I $uint<10> := 0
                    $merge $entry loopback
                       $phi s := 0 $on $entry  s1 $on loopback
                    $endmerge
                    $if (I < mp) $then
                       I := (I+1)
                       s1 := ($mux (I == 0)  0 (s + mem[I]))
                       $place [loopback]
                    $endif
               } 
               $branchblock[sb2]{
                    $declare $storage  J $uint<10> := 0
                    J := (mp + 1)
                    $merge $entry loopback
                       $phi s := 0 $on $entry s1 $on loopback
                    $endmerge
                    $if (J < high) $then
                         J := (J+1)
                         s1 := ($mux (J == (mp+1))  0 (s + mem[J]))
                         $place [loopback]
                    $endif
               } 
           }

           // combine results from parallel statement above
           snontriv := (%parsum%sb1:s  + %parsum%sb2:s)
           $place [nontrivsum]
      $else

           // summation is trivial 
           striv := mem[low]
           $place [trivsum]
      $endif
      $merge nontrivsum trivsum
         // which sum do you pick? depends on which path was taken
         $phi sum := snontriv $on nontrivsum strivsum $on trivsum
      $endmerge
    }

    // bring sum up to the surface.
    sum := %trivcheck:sum
}
\end{verbatim}
This example describes an algorithm which computes the sum of
a section of an array by dividing the problem into two partial
summations.

\section{Syntax} \label{sec:Syntax}

The syntax for \Aa follows the following
principles
\begin{itemize}
\item All keywords begin with the \$ sign.
\item The region between \{ and \} defines a new scope.
\item Statements are space separated (no semicolons at all).
\item Expressions are fully parenthitized.  Thus
$(a + b)$ is a legal expression, but $a+b$ is not.
\end{itemize}

The parser is implemented using 
an LL(k) parser (written as rules to be parsed by antlr2 \cite{ref:antlr2}).
The grammar for the parser is (using the EBNF notation) given below. 
\input{AaParser}

This is the set of tokens recognized by the lexical analyzer (or lexer).
\input{AaLexer}

\begin{thebibliography}{99}
\bibitem{ref:SameerPhD}
Sameer D. Sahasrabuddhe,
``A competitive pathway from high-level programs to hardware,''
Ph.D. thesis, IIT Bombay, 2009.

\bibitem{ref:antlr2}
http://www.antlr2.org.
\end{thebibliography}
